\pkg{cv}: An \proglang{R} Package for Cross-Validation of Regression Models

### 2024-01-04 scratch/notes_john/JSS-article/JSS-article-reduced.Rmd
put plots for contrived mixed-models data in single figure; text needs adjustment

```{r plot1}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Hierarchical data set, showing the first 10 of 100 students."
library("lattice")
library("latticeExtra")
plot <- xyplot(y ~ x, data=Data[1:Nx, ], group=group,
               ylim=c(4, 16),
               par.settings=list(superpose.symbol=list(pch=1, cex=0.7))) +
    layer(panel.ellipse(..., center.cex=0))
plot # display graph
```

@@

```{r}
Data <- within(Data, {
  fit_mod0.fe <- predict(mod.0, re.form = ~ 0) # fixed effects only
  fit_mod0.re <- predict(mod.0) # fixed and random effects (BLUPs)
  fit_mod1.fe <- predict(mod.1, re.form = ~ 0)
  fit_mod1.re <- predict(mod.1)
  fit_mod2.fe <- predict(mod.2, re.form = ~ 0)
  fit_mod2.re <- predict(mod.2)
  fit_mod3.fe <- predict(mod.3, re.form = ~ 0)
  fit_mod3.re <- predict(mod.3)
})
```

@@

```{r}
Data_long <- reshape(Data[1:Nx, ], direction = "long", sep = ".", 
              timevar = "effect", varying = grep("\\.", names(Data[1:Nx, ])))
Data_long$id <- 1:nrow(Data_long)
Data_long <- reshape(Data_long, direction = "long", sep = "_", 
              timevar = "modelcode",  varying = grep("_", names(Data_long)))
Data_long$model <- factor(
  c("~ 1", "~ 1 + x", "~ 1 + x + xm", "~ 1 + I(x - xm)")
  [match(Data_long$modelcode, c("mod0", "mod1", "mod2", "mod3"))]
)
```

@@

```{r plot-fits-mod0}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Predictions from the random intercept model."
(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" &  effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  main="Model: y ~ 1 + (1 | group)",
  key=list(
    corner=c(0.05, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))
```

@@

```{r plot-fits-mod1}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Predictions from the model with random intercepts and $x$ as a fixed-effect predictor."
(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  main="Model: y ~ 1 + x + (1 | group)",
  ylim=c(-15, 35),
  key=list(
    corner=c(0.95, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))
```

@@

Predictions from model \code{mod.2} appear in the following graph:
```{r plot-fits-mod2}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Predictors from the model with random intercepts, $x$, and the group (student) mean of $x$ as predictors."
(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  main="Model: y ~ 1 + x + xm + (1 | group)",
  ylim=c(4, 16),
  key=list(
    corner=c(0.05, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))
```

@@

```{r plot-fits-mod3}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Predictions from the estimated model generating the data."
(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod3" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod3" & effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  main="Model: y ~ 1 + I(x - xm) + (1 | group)",
  ylim=c(4, 16),
  key=list(
    corner=c(0.05, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))
```

>>>>

```{r plot1, echo=FALSE, results=FALSE}
#| fig.height = 4,
#| fig.cap = "(a) Hierarchical data set, showing the first 10 of 100 students, and (b)--(e) several mixed models fit to the data",
#| fig.subcap=rep("", 5),
#| fig.ncol = 2, out.width = "45%", fig.align = "center"
library("lattice")
library("latticeExtra")
library("lme4", quietly=TRUE)
plot <- xyplot(y ~ x, data=Data[1:Nx, ], group=group,
               ylim=c(4, 16),
               par.settings=list(superpose.symbol=list(pch=1, cex=0.7))) +
    layer(panel.ellipse(..., center.cex=0))
plot # display graph

mod.0 <- lmer(y ~ 1 + (1 | group), Data)
mod.1 <- lmer(y ~ x + (1 | group), Data)
mod.2 <- lmer(y ~ x + xm + (1 | group), Data)
mod.3 <- lmer(y ~ I(x - xm) + (1 | group), Data)

Data <- within(Data, {
  fit_mod0.fe <- predict(mod.0, re.form = ~ 0) # fixed effects only
  fit_mod0.re <- predict(mod.0) # fixed and random effects (BLUPs)
  fit_mod1.fe <- predict(mod.1, re.form = ~ 0)
  fit_mod1.re <- predict(mod.1)
  fit_mod2.fe <- predict(mod.2, re.form = ~ 0)
  fit_mod2.re <- predict(mod.2)
  fit_mod3.fe <- predict(mod.3, re.form = ~ 0)
  fit_mod3.re <- predict(mod.3)
})

Data_long <- reshape(Data[1:Nx, ], direction = "long", sep = ".", 
              timevar = "effect", varying = grep("\\.", names(Data[1:Nx, ])))
Data_long$id <- 1:nrow(Data_long)
Data_long <- reshape(Data_long, direction = "long", sep = "_", 
              timevar = "modelcode",  varying = grep("_", names(Data_long)))
Data_long$model <- factor(
  c("~ 1", "~ 1 + x", "~ 1 + x + xm", "~ 1 + I(x - xm)")
  [match(Data_long$modelcode, c("mod0", "mod1", "mod2", "mod3"))]
)

(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" &  effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  main="Model: y ~ 1 + (1 | group)",
  key=list(
    corner=c(0.05, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))

(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  main="Model: y ~ 1 + x + (1 | group)",
  ylim=c(-15, 35),
  key=list(
    corner=c(0.95, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))

(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  main="Model: y ~ 1 + x + xm + (1 | group)",
  ylim=c(4, 16),
  key=list(
    corner=c(0.05, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))

(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod3" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod3" & effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  main="Model: y ~ 1 + I(x - xm) + (1 | group)",
  ylim=c(4, 16),
  key=list(
    corner=c(0.05, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))

```

@@

@@

@@

@@

@@

Predictions from model \code{mod.2} appear in the following graph:

@@

### 2024-01-04 scratch/notes_john/JSS-article/JSS-article-reduced.tex
put plots for contrived mixed-models data in single figure; text needs adjustment

Residual standard error: 4.374 on 389 degrees of freedom
Multiple R-squared:  0.6876,    Adjusted R-squared:  0.686 
F-statistic:   428 on 2 and 389 DF,  p-value: < 2.2e-16
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.auto)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 886597
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
method: Woodbury
criterion: mse
cross-validation criterion = 19.41669
bias-adjusted cross-validation criterion = 19.39385
full-sample criterion = 18.98477 
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> library("lattice")
R> library("latticeExtra")
R> plot <- xyplot(y ~ x, data=Data[1:Nx, ], group=group,
+                ylim=c(4, 16),
+                par.settings=list(superpose.symbol=list(pch=1, cex=0.7))) +
+     layer(panel.ellipse(..., center.cex=0))
R> plot # display graph
\end{CodeInput}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/plot1-1}

@@

\caption[Hierarchical data set, showing the first 10 of 100 students]{Hierarchical data set, showing the first 10 of 100 students.}\label{fig:plot1}
\end{figure}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> Data <- within(Data, {
+   fit_mod0.fe <- predict(mod.0, re.form = ~ 0) # fixed effects only
+   fit_mod0.re <- predict(mod.0) # fixed and random effects (BLUPs)
+   fit_mod1.fe <- predict(mod.1, re.form = ~ 0)
+   fit_mod1.re <- predict(mod.1)
+   fit_mod2.fe <- predict(mod.2, re.form = ~ 0)
+   fit_mod2.re <- predict(mod.2)
+   fit_mod3.fe <- predict(mod.3, re.form = ~ 0)
+   fit_mod3.re <- predict(mod.3)
+ })
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> Data_long <- reshape(Data[1:Nx, ], direction = "long", sep = ".", 
+               timevar = "effect", varying = grep("\\.", names(Data[1:Nx, ])))
R> Data_long$id <- 1:nrow(Data_long)
R> Data_long <- reshape(Data_long, direction = "long", sep = "_", 
+               timevar = "modelcode",  varying = grep("_", names(Data_long)))
R> Data_long$model <- factor(
+   c("~ 1", "~ 1 + x", "~ 1 + x + xm", "~ 1 + I(x - xm)")
+   [match(Data_long$modelcode, c("mod0", "mod1", "mod2", "mod3"))]
+ )
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> (plot +
+   xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" & effect == "fe"),
+          groups=group, type="l", lwd=2) +
+   xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" &  effect == "re"),
+          groups=group, type="l", lwd=2, lty=3)
+ ) |> update(
+   main="Model: y ~ 1 + (1 | group)",
+   key=list(
+     corner=c(0.05, 0.05),
+     text=list(c("fixed effects only","fixed and random")),
+     lines=list(lty=c(1, 3))))
\end{CodeInput}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/plot-fits-mod0-1}

\caption[Predictions from the random intercept model]{Predictions from the random intercept model.}\label{fig:plot-fits-mod0}
\end{figure}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> (plot +
+   xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & effect == "fe"),
+          groups=group, type="l", lwd=2) +
+   xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & effect == "re"),
+          groups=group, type="l", lwd=2, lty=3)
+ ) |> update(
+   main="Model: y ~ 1 + x + (1 | group)",
+   ylim=c(-15, 35),
+   key=list(
+     corner=c(0.95, 0.05),
+     text=list(c("fixed effects only","fixed and random")),
+     lines=list(lty=c(1, 3))))
\end{CodeInput}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/plot-fits-mod1-1}

\caption[Predictions from the model with random intercepts and $x$ as a fixed-effect predictor]{Predictions from the model with random intercepts and $x$ as a fixed-effect predictor.}\label{fig:plot-fits-mod1}
\end{figure}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> (plot +
+   xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & effect == "fe"),
+          groups=group, type="l", lwd=2) +
+   xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & effect == "re"),
+          groups=group, type="l", lwd=2, lty=3)
+ ) |> update(
+   main="Model: y ~ 1 + x + xm + (1 | group)",
+   ylim=c(4, 16),
+   key=list(
+     corner=c(0.05, 0.05),
+     text=list(c("fixed effects only","fixed and random")),
+     lines=list(lty=c(1, 3))))
\end{CodeInput}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/plot-fits-mod2-1}

\caption[Predictors from the model with random intercepts, $x$, and the group (student) mean of $x$ as predictors]{Predictors from the model with random intercepts, $x$, and the group (student) mean of $x$ as predictors.}\label{fig:plot-fits-mod2}
\end{figure}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> (plot +
+   xyplot(fit ~ x, subset(Data_long, modelcode == "mod3" & effect == "fe"),
+          groups=group, type="l", lwd=2) +
+   xyplot(fit ~ x, subset(Data_long, modelcode == "mod3" & effect == "re"),
+          groups=group, type="l", lwd=2, lty=3)
+ ) |> update(
+   main="Model: y ~ 1 + I(x - xm) + (1 | group)",
+   ylim=c(4, 16),
+   key=list(
+     corner=c(0.05, 0.05),
+     text=list(c("fixed effects only","fixed and random")),
+     lines=list(lty=c(1, 3))))
\end{CodeInput}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/plot-fits-mod3-1}

\caption[Predictions from the estimated model generating the data]{Predictions from the estimated model generating the data.}\label{fig:plot-fits-mod3}
\end{figure}
\end{CodeChunk}

>>>>

Residual standard error: 4.374 on 389 degrees of freedom
Multiple R-squared:  0.6876,    Adjusted R-squared:  0.686 
F-statistic:   428 on 2 and 389 DF,  p-value: < 2.2e-16
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.auto)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 838816
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
method: Woodbury
criterion: mse
cross-validation criterion = 19.36368
bias-adjusted cross-validation criterion = 19.34363
full-sample criterion = 18.98477 
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{figure}

{\centering \subfloat[\label{fig:plot1-1}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced_files/figure-latex/plot1-1} }\subfloat[\label{fig:plot1-2}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced_files/figure-latex/plot1-2} }\newline\subfloat[\label{fig:plot1-3}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced_files/figure-latex/plot1-3} }\subfloat[\label{fig:plot1-4}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced_files/figure-latex/plot1-4} }\newline\subfloat[\label{fig:plot1-5}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced_files/figure-latex/plot1-5} }

@@

\caption[(a) Hierarchical data set, showing the first 10 of 100 students, and (b)--(e) several mixed models fit to the data]{(a) Hierarchical data set, showing the first 10 of 100 students, and (b)--(e) several mixed models fit to the data}\label{fig:plot1}
\end{figure}
\end{CodeChunk}

@@

@@

@@

@@

@@

@@
